Synthesizing Symmetric Lenses

### 2019-02-19 papers/SymmetricLenses/main.tex
fixes

\subsection{Simple Symmetric Lenses}
Semantically, a simple symmetric lens between sets $X$ and
$Y$ comprises four functions subject to four
``round-tripping'' laws:

%\begin{enumerate}
%\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
%\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
%\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
%\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
%\end{enumerate}
\begin{center}
\begin{minipage}[c]{.3\textwidth}
\begin{gather}
 \mbox{$\CreateR \OfType X \rightarrow Y$} \nonumber \\
 \mbox{$\CreateL \OfType Y \rightarrow X$} \nonumber \\
 \mbox{$\PutR \OfType X \rightarrow Y \rightarrow Y$} \nonumber \\
 \mbox{$\PutL \OfType Y \rightarrow X \rightarrow X$} \nonumber
\end{gather} 
\end{minipage} 
\hspace{.4in}
\begin{minipage}[c]{.5\textwidth}
\begin{align}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{align}
\end{minipage} 
\end{center}
\noindent
\vskip -.5em
The two \KW{create} functions are used to propagate shared information and fill
in default values when introducing new data (\EG, on \KW{create} the ``unk''
salary entry is added alongside the name to the management file when HR inserts
a new employee). The two \KW{put} functions propagate edits from one format to
the other by combining a new value from one with an old value from the other.
The notation $\Lens.\PutR$ extracts the \PutR function from simple symmetric
lens $\Lens$; similarly for \PutL. These four functions can be used to keep the
common information between two file types in synch. For example, if a new file
of the left-hand format is created, the \CreateR function will a new file in the
right-hand format. If the right-hand format is edited, the \PutL function can
update the left-hand format with the changed information.

The round-tripping laws guarantee that pushing an ``un-edited'' value from one
side (the result of a put or create) back through a lens in the other direction
will get back to exactly where we began. For example, consider,
\lstinline{emp_lens}, a lens between
\lstinline{emp_salaries} and
\lstinline{emp_insurance}. Applying
\lstinline{emp_lens}.\CreateR to a salary file creates a insurance file with the
same employee names in the file, with \lstinline{"UNK"} for each insurance
company. The round-tripping laws guarantee that using \lstinline{emp_lens}.\PutL
to put the generated insurance file into the original salary file will return
the original salary file, unmodified. If \KW{put}ting a string $s$ into a string
$t$ results in the original string $s$, then $s$ and $t$ are considered
\emph{synchronized}.

@@

In principle, programmers can define simple symmetric lenses just by writing
four arbitrary functions by hand and showing they satisfy the round-tripping
laws. But a better idea is to streamline the process by providing a set of
combinators---a domain-specific language---for building law-abiding complex
lenses from simpler ones. Such bidirectional languages have been given for
building lenses over many different kinds of data, including database
relations~\cite{BohannonPierceVaughan} algebraic data
types~\cite{symmetric-lenses}, graphs~\cite{tgg}, and more.

We focus here on combinators for defining lenses between string data formats
described by regular expressions.

If $\BRegex$ and $\BRegexAlt$ are regular expressions, then
\lstinline{$\Lens$ : $\BRegex \Leftrightarrow \BRegexAlt$} indicates that
$\Lens$ is a simple symmetric lens between $\LanguageOf{\BRegex}$ and
$\LanguageOf{\BRegexAlt}$. (We will use undecorated variables later for
stochastic regular expressions; we mark plain REs with overbars.)  When
clear from context, we shorten ``simple symmetric lens'' to just
``lens.''

We illustrate some of these
combinators by defining a lens between the two employee data formats
in Figure~\ref{fig:minimized-representations}. The finished lens, which appears
in Figure~\ref{fig:example_lens}, has type
\lstinline{emp_salaries} $\Leftrightarrow$
\lstinline{emp_insurance}.

The simplest combinator is the identity lens \IdentityLens, which takes as an
argument a regular expression $\BRegex$ and performs an identity mapping on data
matching $\BRegex$. 
%
\begin{lstlisting}
id(name) : name $\Leftrightarrow$ name
\end{lstlisting}
%
The identity lens fully connects the corresponding data in the two formats:
when one is updated, the other gets the exact same value. Both the \CreateR
and \CreateL functions are the identity function ($\CreateR \App s = s$),
and the put functions merely return the first argument
($\PutROf{\String}{\StringAlt} = \String$).
%
In our example, when a \lstinline{name} is updated in one file, the identity
lens ensures it is also updated in the other file.

In contrast, the \Disconnect lens does not propagate any edits at all
from one format to the other.  It takes four arguments: two
regular expressions and two strings. The regular expressions specify the formats
on the two sides, while the strings provide default values. 
%
\begin{lstlisting}
disconnect(salary, "", "unk", "") : salary $\Leftrightarrow$ ""
\end{lstlisting}
%
On creates, the input values are thrown away, and default values are
returned ($\CreateLOf{\StringAlt} =
\lstinline{"unk"}$), and on puts, the second argument is used and the first is
thrown away ($\PutROf{\String}{\StringAlt} = \StringAlt$). In our extended
example, the \lstinline{salary} field is only present in management files,
so we use the disconnect lens to ensure salary edits do not cause updates to the
HR file.
For example, with the above lens, $\PutL \App \lstinline{""} \App 60000$ will return 60000,
and \PutR will always return \lstinline{""}.

@@

Finally, there are a number of combinators that construct
more complex lenses from simpler ones. These include concatenation
($\ConcatLensOf{\Lens_1}{\Lens_2}$, often written $\Lens_1 . \Lens_2$), variation
($\OrLensOf{\Lens_1}{\Lens_2}$), and iteration ($\IterateLensOf{\Lens}$).
For example, the composite lens \lstinline{name_lens} in the running
example
%
\begin{lstlisting}
id(name) . id(" ") . id(name) . del(": ") . ins(",")
\end{lstlisting}
%
will transform ``\lstinline{Jane Doe: }'' to ``\lstinline{Jane Doe,}'' in
the left-to-right direction.

% concat(ins(header),employees_lens) : emp_salaries $\Leftrightarrow$ emp_insurance
%
%\ksf{I don't think this example is necessary.}
%Data can also appear in multiple forms. For example,
%a list of employees with insurance information can either be empty or
%be a sequence of employee/insurance records: \ksf{This example is
%vacuous as you could eliminate the or in favor of the *.  Do somethign
%else to introduce or?}
%
%\begin{lstlisting}
%or(id(""),employees_lens) : emp_salaries $\Leftrightarrow$ "" | ("\n" emp_ins) . ("\n" emp_ins)*
%\end{lstlisting}
The \IterateLens lens also appears in our running example, where it
synchronizes data for a list of employees when given a
lens \lstinline{employee_lens} that synchronizes  data for a single employee:
%
\begin{lstlisting}
iterate(id("\n") . employee_lens) : ("\n" . emp_salary)$^*$ $\Leftrightarrow$ ("\n" . emp_ins)$^*$
\end{lstlisting}
%

@@

We provide typing rules for the simple symmetric combinators, most of
them syntax directed. However, regular expressions have a number of
equivalences on them, and sometimes we must convert one regular expression type
to another, equivalent, regular expression type.
%
\[
  \centering
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \BRegex \SSREquiv \BRegex'\\
    \BRegexAlt \SSREquiv \BRegexAlt'
  }
  {
    \Lens \OfType \BRegex' \Leftrightarrow \BRegexAlt'
  }
\]
%
This rule is needed for lenses like the final one in
Figure~\ref{fig:example_lens} to be well-typed.

@@

\subsection{Synthesizing Simple Symmetric Lenses}

Lens expressions for small examples like the one above are pretty easy to
write.  But for larger and more complex formats, it can become quite difficult to
write lenses by hand.  Instead, we wish to synthesize them from examples.

That is, given a pair of regular expressions and a set of input-output examples, we want
to find a simple symmetric lens that is typed by the regular expression pair and
satisfies all the input/output examples. We call such a lens a \emph{satisfying
  lens}. In our running example, we wish to synthesize a lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance}, using as an input-output
example the data in Figure~\ref{fig:minimized-representations}.

One challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries, emp_insurance, rep(emp_salaries), rep(emp_insurance))
\end{lstlisting}
%
is another well-typed lens (where \lstinline{rep($\BRegex$)} denotes some string in the
language of $\BRegex$).  Adding the input-output example in
Figure~\ref{fig:minimized-representations} 
helps, but only a little: In general, {\em many} examples may be required to rule
out all possible occurrences of nested \Disconnect lenses,
particularly in complex formats. 
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
While we could ask programmers to provide detailed and precise specifications
for their synthesis tasks, we would prefer an algorithm that can intelligently
choose a lens despite the underspecification. Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

How can we identify such a ``likely'' lens? We propose the following heuristic:
A satisfying lens is ``more likely'' if it ``connects'' more data between the
two formats, \EG, by using an identity rather than a disconnect lens. The
likeliest satisfying lenses would connect as much data as possible. More
formally, we define the likelihood of a satisfying lens as the expected number
of bits required to recover data in one format from synchronized data in the
other. Higher likelihoods correspond to fewer bits. Yet more formally, we say
that two strings $s$ and $t$ are \emph{synchronized according to lens $l$} if
$l.\PutR\ s\ t = t$ and $l.\PutL\ t\ s = s$. We say that we can \emph{recover}
$s$ from $t$ using bits $b$ and lens $l$ if we can reconstruct $s$ from $t$,
$b$, and $l$. For example, given the \IdentityLens{} lens, we can recover $s$
from $t$ using no bits because, in this case, $s$ is just $t$. In contrast,
given the \Disconnect{} lens, we need enough bits to fully encode $s$ in order
to recover it from $t$ because $t$ contains no useful information. Thus, if both
\IdentityLens{} and \Disconnect{} are satisfying lenses for a particular pair of
formats, then \IdentityLens{} will be more likely.

The expected number of bits required to recover a piece of data corresponds to
the well-known information-theoretic concept of
\emph{entropy}~\cite{Shannon1948}. Calculating entropy requires a probability
distribution over the space of possible values for the data. Specifically, given
a set $S$ and a probability distribution $P \OfType S \rightarrow \Reals$ over
$S$, the entropy $\EntropyOf{S,P}$ is $-\Sigma_{s\in S}P(s)\Log_2P(s)$. In
information theory, each element of $S$ has \emph{information content}, the
number of bits required to specify that element. The information content of an
element $s$ is $-\Log_2P(s)$. For example, let $S = \SetOf{a,b_1,b_2}$, where
$P(a)=.5$ and $P(b_1) = .25 = P(b_2)$. The information content of $a$ is $1$,
and the information content of each $b_i$ is 2. The entropy of $S$ calculates
the expected number of bits required to specify a random element pulled from
$S$, in other words, the probability of each element times its information
content. In our example, the entropy of $S$ is $1.5$. When $P$ is clear, we will
often use the shorthand of $\EntropyOf{S}$ for $\EntropyOf{S,P}$.

In our setting, we already have a way of expressing sets of data:
regular expressions.  What we need is a way to express probability
distributions over that data.
To that end, we adopt \emph{stochastic regular expressions}~\cite{stoch-def,stoch-rnn} (SREs), 
which are regular expressions in which each operator is annotated with a
probability (see \S\ref{sec:sre}).  A stochastic regular expression
specifies both a set of strings and a probability
distribution over that set, enabling us to calculate the
entropy of the SRE.

In the synthesis setting, there are typically an infinite number of possible
lenses that satisfy the specification. There are many choices on which
components of the data should be synchronized vs which should be left
unsynchronized. We use entropy to guage the relative likelihood of such lenses
in our synthesis algorithm (see \S\ref{subsec:lenscosts}). For any lens $\Lens$, we can calculate the expected
number of bits required to recover a string $t$ in $\LanguageOf{\RegexAlt}$ from
a synchronized string $s$ in $\LanguageOf{\Regex}$. This expectation is the
\emph{conditional entropy of $\Regex$ given $\RegexAlt$ and $\Lens$}, formally
$\sum_{s \in \Regex}\ProbabilityOf{\Regex}{s}\cdot\EntropyOf{\SetOf{t \SuchThat
    \Lens.\PutROf{s}{t} = t}}$. The likelihood we assign to $\Lens$ is the sum
of the conditional entropy of $\Regex$ given $\RegexAlt$ and $\Lens$ and the
conditional entropy of $\RegexAlt$ given $\Regex$ and $\Lens$. This metric
assigns higher entropy (or cost) to lenses where knowing the string on one side
provides little information about the string on the other side. It assigns zero
entropy to bijections because given a string $s \in \Regex$, the bijection
exactly determines the corresponding string in $\RegexAlt$.

@@

To obtain SREs from the plain regular expressions that users write, we use a
heuristic that attempts to assign probability annotations that give each
string in the language equal probablity. Our synthesis algorithm could
equally well take as input a pair of stochastic regular expression rather than a
plain one, permitting users to specify a data-specific probability
distribution (either manually calculated or inferred from some dataset of
interest). However, we have found that our heuristic works well in practice,
so we have not explored these options.

Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work~\cite{augustsson-2004, osera+:pldi15,
  feser-pldi-2015, frankle+:popl16} has shown that we can use types to
dramatically restrict the search space and improve the effectiveness of
synthesis. In our setting, types are pairs of stochastic regular
expressions. The fact that each regular expression is semantically equivalent to
infinitely many others complicates the problem.
Following existing work on bidirectional program
synthesis~\cite{maina+:quotient-synthesis}, we 
split our algorithm into two communicating
searches. The first search,
\Expand, navigates the space of semantically equivalent regular expressions
by applying rewrite rules that preserve both the semantics and the
probability distributions of those stochastic regular expressions. This
algorithm ranks pairs of stochastic regular expressions by the number of
rewrite rule applications required to obtain each pair from the one given as
input. It passes the pairs off to the second 
search procedure, \GreedySynth, in rank order with the smallest first.

Abstractly, \GreedySynth looks for highest-likelihood lenses between a given
pair of stochastic regular expressions $\Regex$ and $\RegexAlt$ by performing a
type-directed search. In more detail, \GreedySynth first normalizes the
stochastic regular expressions provided by \Expand into \emph{stochastic DNF
  regular expressions}---a constrained representation of stochastic regular
expressions with disjunctions distributed over concatenations and with
concatenations and disjunctions normalized to operating over lists (see \S\ref{subsec:sdnfre}).
\GreedySynth then uses the syntax of these n-ary DNF regular expressions to find
normalized lenses, in a form we call \emph{simple symmetric n-ary DNF lenses}.
These lenses include neither a composition operator nor a type equivalence rule.
These restrictions mean that each simple symmetric n-ary DNF lens is typed by a
single pair of stochastic n-ary DNF regular expressions, hence \GreedySynth's
search space for a given pair of regular expressions is finite. The variations
that \GreedySynth considers include choosing between \IdentityLens{} and
\Disconnect lenses and deciding how to match multiple occurrences of the same
kind of data, \EG, the two occurrences of \lstinline{name} in the management and
HR data formats (should the algorithm match the first instance of
\lstinline{name} on the left with the first instance of \lstinline{name} on the
right, or to the second). Finally, \GreedySynth yields a simple symmetric lens
by converting the n-ary forms into the binary forms provided in the surface
language.

To illustrate the partnership between \Expand and \GreedySynth, consider
synthesizing the lens shown in Figure~\ref{fig:example_lens}. When \GreedySynth
initially tries to find a lens between \lstinline{emp_salaries} and
\lstinline{emp_insurance}, it hits some snags: there are complex disjunctions
and concatenations in \lstinline{emp_salaries} outside the iteration, but no
such complexity in \lstinline{emp_insurance}. Because of this disparity between
the formats, \GreedySynth finds an unlikely lens (involving undesired
\KW{merge}s and \KW{disconnect}s), and the search continues. However, after
\Expand applies a star-semiring axiom (presented in Figure~\ref{fig:ssr-sre}) to
rewrite the regular expression \lstinline{("\n" . emp_salary)$^*$} within
\lstinline{emp_salaries} into
\lstinline{"" | (("\n" . emp_salary) . ("\n" . emp_salary)$^*$)}, the
subcomponents line up better, and the more likely correct lens is instead found.

This architecture of communicating synthesizers gives us a way to enumerate
pairs of stochastic regular expressions of increasing rank and to efficiently
search through them, but it poses a problem: when should \Expand stop proposing
new SRE pairs? We may have found a promising lens between a pair of stochastic
regular expressions, but a different pair we haven't yet discovered may give
rise to an even better lens.

>>>>

\subsection{Simple Symmetric Lenses}
Semantically, a simple symmetric lens (or lens when clear from context) between
sets $X$ and $Y$ comprises four functions subject to four ``round-tripping''
laws.

%\begin{enumerate}
%\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
%\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
%\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
%\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
%\end{enumerate}
\begin{center}
\begin{minipage}[c]{.3\textwidth}
\begin{gather}
 \mbox{$\CreateR \OfType X \rightarrow Y$} \nonumber \\
 \mbox{$\CreateL \OfType Y \rightarrow X$} \nonumber \\
 \mbox{$\PutR \OfType X \rightarrow Y \rightarrow Y$} \nonumber \\
 \mbox{$\PutL \OfType Y \rightarrow X \rightarrow X$} \nonumber
\end{gather} 
\end{minipage} 
\hspace{.4in}
\begin{minipage}[c]{.5\textwidth}
\begin{align}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{align}
\end{minipage} 
\end{center}
\noindent
\vskip -.5em
The two \KW{create} functions are used to propagate shared information and fill
in default values when introducing new data (\EG, on \KW{create} the ``unk''
salary entry is added alongside the name to the management file when HR inserts
a new employee). The two \KW{put} functions propagate edits from one format to
the other by combining a new value from one with an old value from the other.
The notation $\Lens.\PutR$ extracts the \PutR function from the
lens $\Lens$; similarly for \PutL. These four functions can be used to keep the
common information between two file types in synch. For example, if a new file
of the left-hand format is created, the \CreateR function will a new file in the
right-hand format. If the right-hand format is edited, the \PutL function can
update the left-hand format with the changed information.

The round-tripping laws guarantee that pushing an ``un-edited'' value from one
side (the result of a put or create) back through a lens in the other direction
will get back to exactly where we began. For example, consider a simple
symmetric lens
synchronizing the employee data formats in
Figure~\ref{fig:minimized-representations}. Applying the \CreateR function to a
salary file creates a insurance file with the same employee names in the file,
with
\lstinline{"UNK"} for each insurance company. The round-tripping
laws guarantees that applying the \PutL function to the generated insurance file
and the original salary file will return the original salary file, unmodified. If \KW{put}ting a string $s$ into a string $t$ results in the original string $s$, then $s$ and $t$ are considered \emph{synchronized}.

@@

In principle, programmers can define simple symmetric lenses just by writing
four arbitrary functions by hand and showing they satisfy the round-tripping
laws. But a better idea is to streamline the process by providing a set of
combinators---a domain-specific language---for building complex, law-abiding
lenses from simpler ones. Such bidirectional languages have been given for
building lenses over many different kinds of data, including database
relations~\cite{BohannonPierceVaughan} algebraic data
types~\cite{symmetric-lenses}, graphs~\cite{tgg}, and more.

We focus here on combinators for defining lenses between string data formats
described by regular expressions.
If $\BRegex$ and $\BRegexAlt$ are regular expressions, then
\lstinline{$\Lens$ : $\BRegex \Leftrightarrow \BRegexAlt$} indicates that
$\Lens$ is a simple symmetric lens between $\LanguageOf{\BRegex}$ and
$\LanguageOf{\BRegexAlt}$. (We will use undecorated variables later for
stochastic regular expressions; we mark plain REs with overbars.)
We illustrate some of these
combinators by defining lenses on subcomponents of the employee data formats.

The simplest combinator is the identity lens \IdentityLens, which takes as an
argument a regular expression $\BRegex$ and performs an identity mapping on data
matching $\BRegex$. 
%
\begin{lstlisting}
id(name) : name $\Leftrightarrow$ name
\end{lstlisting}
%
The identity lens moves data back and forth from source to target without
changing it. Both the \CreateR and \CreateL functions are the identity function
($\CreateR \App s = s$), and the put functions merely return the first argument
($\PutROf{\String}{\StringAlt} = \String$).

In contrast, $\DisconnectOf{\Regex}{\RegexAlt}{\String}{\StringAlt}$ does not propagate any data at all
from one format to the other.  The \Disconnect lens takes four arguments: two
regular expressions ($\Regex, \RegexAlt$) and two strings ($\String, \StringAlt$). The regular expressions specify the formats
on the two sides, while the strings provide default values. 
%
\begin{lstlisting}
disconnect(salary, "", "unk", "") : salary $\Leftrightarrow$ ""
\end{lstlisting}
%
On creates, the input values are thrown away, and default values are
returned ($\CreateLOf{\StringAlt} =
\lstinline{"unk"}$), and on puts, the second argument is used and the first is
thrown away ($\PutROf{\String}{\StringAlt} = \StringAlt$). For 
example, the \lstinline{salary} field is only present in management files,
so the disconnect lens can ensure salary edits do not cause updates to the
HR file.
For example, with the above lens, $\PutL \App \lstinline{""} \App 60000$ will return 60000,
and \PutR will always return \lstinline{""}.

@@

Finally, there are a number of combinators that construct
more complex lenses from simpler ones. These include concatenation
($\ConcatLensOf{\Lens_1}{\Lens_2}$, often written $\Lens_1 . \Lens_2$), variation
($\OrLensOf{\Lens_1}{\Lens_2}$), and iteration ($\IterateLensOf{\Lens}$).
For example, we could use the following lens
%
\begin{lstlisting}
id(name) . id(" ") . id(name) . del(": ") . ins(",")
\end{lstlisting}
%
to transform ``\lstinline{Jane Doe: }'' to ``\lstinline{Jane Doe,}'' in
the left-to-right direction.

% concat(ins(header),employees_lens) : emp_salaries $\Leftrightarrow$ emp_insurance
%
%\ksf{I don't think this example is necessary.}
%Data can also appear in multiple forms. For example,
%a list of employees with insurance information can either be empty or
%be a sequence of employee/insurance records: \ksf{This example is
%vacuous as you could eliminate the or in favor of the *.  Do somethign
%else to introduce or?}
%
%\begin{lstlisting}
%or(id(""),employees_lens) : emp_salaries $\Leftrightarrow$ "" | ("\n" emp_ins) . ("\n" emp_ins)*
%\end{lstlisting}
The \IterateLens lens is useful for synchronizing a series of items or rows in a
table. For example given a
lens \lstinline{employee_lens} that synchronizes data for a single employee the
following lens
%
\begin{lstlisting}
iterate(id("\n") . employee_lens) : ("\n" . emp_salary)$^*$ $\Leftrightarrow$ ("\n" . emp_ins)$^*$
\end{lstlisting}
%
transforms a list of employees in employees in the Management format to a list
of employees in the HR format in the left-to-right direction. These combinators
are combined in figure~\ref{fig:example_lens} to construct a complete lens
between the employee formats.

@@

Most of the typing rules for lenses are syntax driven, where only a handful of
rules result in a given type.  For example, the type of a
\ConcatLens lens is the concatenation of the regular expressions of the sub-lenses.
\[
  \inferrule*
  {
    \Lens_1 \OfType \BRegex_1 \Leftrightarrow \BRegexAlt_1\\
    \Lens_2 \OfType \BRegex_2 \Leftrightarrow \BRegexAlt_2
  }
  {
    \ConcatLensOf{\Lens_1}{\Lens_2} \OfType \BRegex_1 \Concat \BRegex_2
    \Leftrightarrow
    \BRegexAlt_1 \Concat \BRegexAlt_2
  }
\]
With one notable exception, lens typing rules maintain structural similarity
between regular expression types and lens terms; type equivalence does not. Type
equivalence permits a well-typed lens to be well-typed among any equivalent
regular expression pair.
%
\[
  \centering
  \inferrule*
  {
    \Lens \OfType \BRegex \Leftrightarrow \BRegexAlt\\
    \BRegex \SSREquiv \BRegex'\\
    \BRegexAlt \SSREquiv \BRegexAlt'
  }
  {
    \Lens \OfType \BRegex' \Leftrightarrow \BRegexAlt'
  }
\]
%
This rule is needed for lenses like the final one in
Figure~\ref{fig:example_lens} to be well-typed.

@@

\subsection{Ranking Simple Symmetric Lenses}

Even for small examples, lenses like the one above take a little getting used
to. For larger and more complex formats, it can become quite difficult to
write lenses by hand.  Instead, we wish to synthesize them from examples. 
That is, given a pair of regular expressions and a set of input-output examples, we want
to find a simple symmetric lens that is typed by the regular expression pair and
satisfies all the input/output examples. We call such a lens a \emph{satisfying
  lens}. In our running example, we wish to synthesize a lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance}, using as an input-output
example the data in Figure~\ref{fig:minimized-representations}.

One challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disconnect(emp_salaries, emp_insurance, ..., ...)
\end{lstlisting}
%
is another well-typed lens that satisfies the example in
Figure~\ref{fig:minimized-representations} (where \lstinline{...} represents
components of the example itself). In general, {\em many} examples may be
required to rule out all possible occurrences of nested \Disconnect lenses,
particularly in complex formats.
%\ksf{I don't understand the following.  We shouldn't explain the issue in
%terms of \PutR and \PutL as we are focused on the combinators rather
%than the raw functions.}
%The default strings given to \Disconnect merely need to be the strings
%in the input-output examples, and the lens will still satisfy the
%specification.  While the disconnect lens is prohibited by two
%examples, or by a carefully chosen input-output example on \PutR
%or \PutL, this problem continues presenting itself for the sublenses,
%and is highlighted even more on larger formats.
While we could ask programmers to provide detailed and precise specifications
for their synthesis tasks, we would prefer an algorithm that can intelligently
choose a lens despite the underspecification. Instead of merely finding
\emph{any} satisfying lens, we wish to synthesize a satisfying lens that is
likely to please the user.

How can we identify such a ``likely'' lens? We propose the following heuristic:
A satisfying lens is ``more likely'' if it u;ses more data from one format to
construct the other. One manifestation of this principle involves choosing the
identity lens (which uses all the data) rather than the disconnect lens (which
uses none of the data). We define the likelihood of a satisfying lens as the
expected number of bits required to recover data in one format from synchronized
data in the other. Higher likelihoods correspond to fewer bits. We say that two
strings $s$ and $t$ are \emph{synchronized according to lens $l$} if $l.\PutR\
s\ t = t$ and $l.\PutL\ t\ s = s$. We say that we can \emph{recover} $s$ from
$t$ using bits $b$ and lens $l$ if we can reconstruct $s$ from $t$, $b$, and
$l$. For example, given the \IdentityLens{} lens, we can recover $s$ from $t$
using no bits because, in this case, $s$ is just $t$. In contrast, given the
\Disconnect{} lens, we need enough bits to fully encode $s$ in order to recover
it from $t$ because $t$ contains no useful information. Thus, if both
\IdentityLens{} and \Disconnect{} are satisfying lenses for a particular pair of
formats, then \IdentityLens{} will be more likely.

The expected number of bits required to recover a piece of data corresponds to
the well-known information-theoretic concept of
\emph{entropy}~\cite{Shannon1948}. Calculating entropy requires a probability
distribution over the space of possible values for the data. Specifically, given
a set $S$ and a probability distribution $P \OfType S \rightarrow \Reals$ over
$S$, the entropy $\EntropyOf{S,P}$ is $-\Sigma_{s\in S}P(s)\Log_2P(s)$. In
information theory, each element of $s \in S$ has \emph{information content},
the number of bits required to specify that element in a perfect encoding scheme
($-\Log_2P(s)$). The entropy of $S$ calculates the expected number of bits
required to specify a random element pulled from $S$, in other words, the
probability of each element times its information content. Entropy maintains
that if a data source contains many possible elements, where none have
significantly higher probability than others, it will have high information
content. Data sources with few elements that don't have vanishingly small
probabilities have small entropy. When $P$ is clear, we will often use the
shorthand of $\EntropyOf{S}$ for $\EntropyOf{S,P}$.

In our setting, we already have a way of expressing sets of data: regular
expressions. To calculate entropy, what we need is a way to express probability
distributions over that data. To that end, we adopt \emph{stochastic regular
  expressions}~\cite{stoch-def,stoch-rnn} (SREs), which are regular expressions
in which each operator is annotated with a probability (see \S\ref{sec:sre}). A
stochastic regular expression specifies both a set of strings and a probability
distribution over that set.

We use entropy to guage the relative likelihood of such lenses in our synthesis
algorithm (see \S\ref{subsec:lenscosts}). For any lens $\Lens$, we can calculate
the expected number of bits required to recover a string $t$ in
$\LanguageOf{\RegexAlt}$ from a synchronized string $s$ in
$\LanguageOf{\Regex}$. This expectation is the \emph{conditional entropy of
  $\Regex$ given $\RegexAlt$ and $\Lens$}, formally $\sum_{s \in
  \Regex}\ProbabilityOf{\Regex}{s}\cdot\EntropyOf{\SetOf{t \SuchThat
    \Lens.\PutROf{s}{t} = t}}$. The likelihood we assign to $\Lens$ is the sum
of the conditional entropy of $\Regex$ given $\RegexAlt$ and $\Lens$ and the
conditional entropy of $\RegexAlt$ given $\Regex$ and $\Lens$. This metric
assigns higher entropy (or cost) to lenses where knowing the string on one side
provides little information about the string on the other side. It assigns zero
entropy to bijections because given a string $s \in \Regex$, the bijection
exactly determines the corresponding string in $\RegexAlt$.

@@

To obtain SREs from the plain regular expressions that users write, by default
we use a heuristic that attempts to assign probability annotations that give
each string in the language equal probablity. In other words, this heuristic
tries to make each bit of data no more or less likely than any other.
Our synthesis algorithm could equally well take as input a pair of stochastic
regular expression rather than a plain one, permitting users to specify a
data-specific probability distribution (either manually calculated or inferred
from some dataset of interest).

However, sometimes users already know that certain data should or should not be
used to construct the other format. We introduce relevance annotations to SREs
that enable users to specify whether a piece of data should be used to construct
the other format (with \SRequire) or not (with \Skip). In our example,
\lstinline{salary} and \lstinline{company} could safely be skipped (as they are
disconnected), where \lstinline{name} in \lstinline{emp_salary} and
\lstinline{emp_ins} be annotated as required (as they are converted with the
identity lens).

\subsection{Searching for Likely Lenses}
Now that we have a way to calculate likelihood, we need a way to search for
likely lenses. Previous work~\cite{augustsson-2004, osera+:pldi15,
  feser-pldi-2015, frankle+:popl16} has shown that we can use types to
dramatically restrict the search space and improve the effectiveness of
synthesis. In our setting, types are pairs of stochastic regular
expressions. The fact that each regular expression is semantically equivalent to
infinitely many others complicates the problem.
Following existing work on bidirectional program
synthesis~\cite{maina+:quotient-synthesis}, we 
split our algorithm into two communicating
searches. The first search,
\Expand, navigates the space of semantically equivalent regular expressions
by applying rewrite rules that preserve both the semantics and the
probability distributions of those stochastic regular expressions. This
algorithm ranks pairs of stochastic regular expressions by the number of
rewrite rule applications required to obtain each pair from the one given as
input. It passes the pairs off to the second 
search procedure, \GreedySynth, in rank order with the smallest first.

Abstractly, \GreedySynth looks for highest-likelihood lenses between a given
pair of stochastic regular expressions $\Regex$ and $\RegexAlt$ by performing a
type-directed search. In more detail, \GreedySynth first converts the stochastic
regular expressions provided by \Expand into \emph{stochastic DNF regular
  expressions}---a constrained representation of stochastic regular expressions
with disjunctions distributed over concatenations and with concatenations and
disjunctions normalized to operating over lists (see \S\ref{subsec:sdnfre}).
\GreedySynth then uses the syntax of these n-ary DNF regular expressions to find
normalized lenses, in a form we call \emph{simple symmetric n-ary DNF lenses}.
These lenses include neither a composition operator nor a type equivalence rule.
These restrictions mean that there are only a few simple symmetric n-ary DNF
lenses that are well-typed between each pair of stochastic n-ary DNF regular
expressions, hence \GreedySynth's search space for a given pair of regular
expressions is finite. Finally, \GreedySynth yields a simple symmetric lens by
converting the n-ary forms into the binary forms provided in the surface
language.
